Differences in the Risk of Grade Retention for Biracial and Monoracial Students in the US, 2010-2019

### 2022-12-06 products/paper/main.qmd
Reduce word count of manuscript
If submitting to SRE, it needs to be under 10K words total, including abstract, tables, figures, footnotes, and references. it should now be about 100 words below that.

---
title: Grade Retention and Mixed Race
author: 
  - Aaron Gullickson, *University of Oregon*
abstract: This is a test abstract
thanks: The corresponding author can be reached at aarong@uoregon.edu. All code and data for this project are available at ...
format:
  pdf:
    documentclass: scrartcl
    papersize: letter
    fig-height: 4
    fig-pos: t
    keep-tex: false
  docx:
    toc: false
bibliography: ../project.bib
execute: 
  eval: true
  echo: false
  warning: false
  error: false
editor: source
---

@@

The experiences of mixed race individuals will play an important role in the future of race in the US. A rise in interracial marriage and coupling since the 1960s has resulted in a "biracial baby boom" of individuals whose parents cross racial lines [@root_racially_1992] and this demographic surge may have a transformative effect on racial boundaries [@hochschild_creating_2012; @alba_great_2020]. While racial mixing is nothing new to US history [@morning_who_2000; @gullickson_choosing_2011], this biracial boom is notable for its demographic size and has occurred in a post-Civil Rights environment in which historical norms of racial classification may be eroding.

Understanding the transformative potential of this biracial baby boom is contingent on developing a better understanding of both how these biracial individuals identify their own race and how they are sorted into a prevailing system of racial inequality. While extensive scholarship documents the identification and classification decisions made by or on behalf of multiracial individuals [@xie_racial_1997; @qian_options_2004; @roth_end_2005a; @brunsma_interracial_2005; @herman_forced_2004; @holloway_place_2009; @bratter_mother_2009; @davenport_role_2016; @liebler_boundaries_2016], only a smattering of research has documented how the life chances of biracial individuals compare to their monoracial peers [@kao_racial_1999; @campbell_multiracial_2009; @herman_blackwhiteother_2009]. In this article, I expand on this work by examining the risk of grade retention for K-12 students whose parents identify with different races.

By examining this outcome, I am able to overcome two important limitations that have hampered prior work on the life chances of biracial individuals. First, using self identification of biracial respondents to analyze outcomes is problematic. Prior research indicates that individuals whose parents belong to different races do not always consistently identify with both races [@doyle_are_2007; @liebler_america_2017; @gullickson_racial_2019]. Furthermore, social status and social mobility can predict racial identification generally [@saperstein_racial_2012; @saperstein_mulatto_2013] and social status is related to the identification choices of biracial individuals specifically [@roth_end_2005a; @davenport_role_2016]. Thus, differences in outcomes among biracial individuals may in part produce differences in identification, rather than the other way around.

This problem of reverse causation can be largely addressed when researchers have access to the racial identification of a respondent's biological parents. While such information is available on some large scale longitudinal surveys, researchers then run into a second limitation of sample size. The samples of biracial respondents in such data sources are generally far too small to make reliable statistical inferences. This limitation is particularly severe for biracial individuals with two non-white parents.

When children reside with both biological parents, biracial individuals can be identified in large scale data sources such as the Census. Researchers have previously utilized this feature to examine the classification choices made on behalf of these children [@xie_racial_1997; @roth_end_2005a]. Analyzing outcomes for such children is more difficult, for the simple reason that they do not have many measurable outcomes yet. However, one important outcome, grade retention, is measurable and important even for young children. Being held back in school is a relatively common occurrence that has consequences for later educational transitions [@jimerson_metaanalysis_2001] and is disproportionately applied by race [@warren_patterns_2014]. Grade retention is arguably one of the earliest stratifiers of life chances that an individual will experience.

In this article, I use data from the American Community Survey to examine the likelihood of being held back in school for biracial children in comparison to their monoracial peers. Pooling data from 2010 to 2019, I am able to draw a large enough sample to make reliable statistical inferences for most biracial groups, including those with two non-white parents. The results provide a window into the placement of biracial children within an existing system of racial inequality. Given the large sample size I am able to draw, I am able to estimate relatively precise estimates for most biracial groups. In contrast to prevailing hypotheses that predict either a pattern of hypodescent or an in-between position for biracial individuals, I find a consistent pattern in which the risk of grade retention is similar to that of the lower risk constituent monoracial group. This result holds even after adjusting for differences in resources across respondents, despite the fact that most biracial groups are positively selected by parental resources relative to their monoracial group. I discuss the implications for these findings in the conclusions.

@@

I distinguish three mechanisms that help us understand where biracial individuals might be positioned among these possibilities. First, I consider arguments that focus on the effects of racial discrimination at both an individual and systemic level. Second, I consider arguments regarding different distributions of resources across racial groups. Third, I consider arguments regarding how the nature of hybridity itself may affect outcomes.

@@

Discrimination and bias by race play an important role in overall levels of racial inequality [@quillian_new_2006; @pager_sociology_2008]. Discrimination need not be consciously recognized by actors, but may instead reflect cognitive biases and schema that individuals unknowingly rely upon and reinforce in social interactions [@quillian_new_2006]. Such discrimination will be most consequential when performed by important gatekeepers of opportunity such as teachers, health care workers, and employers. Discrimination need not just occur on an individual basis, but can also be systemic in nature when biases are built into and reinforced by organizations and institutional systems [@reskin_race_2012]. In our current "colorblind" regime [@bonilla-silva_racism_2006], such systems may be formally race neutral but produce disparate impacts by race [@pager_sociology_2008].

@@

The strongest historical precedent for understanding such classification in the US is the "one drop rule" norm governing classification as Black. According to the one-drop rule, individuals of known Black ancestry are considered exclusively Black, regardless of other ancestries [@davis_who_1991]. @davis_who_1991 generalizes this practice into the concept of *hypodescent* in which individuals with mixed racial ancestry will be classified as members of the lower status group. By comparison a pattern of *hyperdescent* implies that individuals of mixed racial ancestry will be classified as members of the higher status group. This pattern of hyperdescent has historically governed the experiences of individuals with American Indian ancestry [@snipp_american_1989; @wolfe_land_2001; @gullickson_choosing_2011].

Although the hypodescent/hyperdescent paradigm is often used to theoretically frame the possible contemporary experiences of multiracial individuals, there are reasons to be cautious about the generalizability of this paradigm. Hypodescent and hyperdescent are observed in two specific historical cases involving Black ancestry and American Indian ancestry, respectively. As @iverson_regimes_2022 note, these two cases could just as parsimoniously be explained by a "dominance" model which ranks ancestries by their tendencies to be supercessive or recessive in determining identification and classification. In this framework, within the historical "racial triangle" of American colonialism, Black ancestry is supercessive, Indigenous ancestry is recessive, and White ancestry falls somewhere in between.

The development of these particular norms may also reflect the historical regime in which they were developed [@gullickson_choosing_2011; @iverson_regimes_2022]. In this case, neither the hypodescent/hyperdescent or dominance paradigm may help us to fully understand the experiences of those of Latino and Asian ancestry due to more recent processes of racialization. Furthermore, the historical norms for all groups may be weakening or transforming in the post-Civil Rights era.

If outcomes for biracial individuals all closer to one of their constituent groups, then we may have stronger evidence about such norms of classification. However, what if those outcomes fall in-between the two groups? How might classification and discrimination produce such an in-between status? Several scenarios might produce such an in-between status.

First, racial ambiguity in appearance may lead to more inconsistent classification into single race categories across observers. Even if norms consistently indicate a certain type of classification, correct identification by observers is still a probabilistic phenomenon in all social interactions and likely to be more heterogeneous for biracial individuals. As a result, biracial individuals will experience a lower sum total of discrimination and bias affecting the lower status group in the case of hypodescent and will not reap the same rewards as a member of the higher status group in the case of hyperdescent.

Second, even when "correctly" classified, greater physical ambiguity may allow biracial individuals to avoid the worst instances of bias and discrimination because they do not conform to the ideal type for the group facing the most discrimination. In other words, biracial people's experience of discrimination may be "softened" because their ambiguous physical appearance provokes less racial antipathy. The US already has a long history of this kind of softening. Because of the one-drop rule, Black individuals vary substantially in skin tone and those individuals with a lighter skin tone have better outcomes [@keith_skin_1991; @hughes_significance_1990; @monk_skin_2014] and are viewed more favorably by both Blacks and Whites [@maddox_cognitive_2002a]. Thus, light skin moderates the stigma of Blackness even among monoracial individuals.

Third, it is also possible that biracial individuals may be classified into an ambiguous "middleman" minority position. This kind of middle tier status for individuals of mixed race is common in Latin America [@telles_race_2009] and @bonilla-silva_biracial_2004 has argued that such a system may be emerging in the contemporary US as well. In this case, many biracial groups may occupy a nebulous and ill-defined middleman or buffer class position in-between whites and darker-skinned non-whites, helping to cement a new kind of racial hierarchy.

@@

Racial inequality is also at least partly explained by different distributions of material, cultural, and social resources across racial groups [@conley_being_1999]. On average, socioeconomic background (e.g. parental income, wealth, and education) differs by race and such socioeconomic background is consequential for outcomes. Other less materially defined resources also vary across groups such as cultural styles, social networks, and neighborhood quality. These differences in background resources reflect the accumulation of historical processes of discrimination and, when combined with contemporary processes of discrimination and bias, magnify the overall level of racial inequality we observe.

How do these differential distributions of resources affect biracial individuals? A naive argument would be that because biracial individuals have parents from two different racial groups, they will on average inherit parental resources roughly halfway between their two constituent groups [@chew_american_1989]. Prior research looking at biracial children in the Census data has found some evidence to support this argument [@chew_american_1989]. In such a case, the resources of biracial individuals would predict outcomes in-between the two constituent groups, barring other mechanisms.

However, the actual resource position of biracial individuals is likely far more complicated because people are not randomly selected into interracial unions. The relationship between this selection process and a biracial child's outcomes are complex because it depends on the strength and direction of that selection on each partner, its consistency across different types of unions, and how strongly the selected characteristic predicts outcomes in the next generation.

Status exchange theory predicts that given an unequal racial hierarchy, individuals from a lower status racial group will be positively selected on socioeconomic characteristics into intermarriage with a higher status group, while individuals from a higher status racial group will be similarly negatively selected into the same marriage [@davis_intermarriage_1941; @merton_intermarriage_1941; @fu_racial_2001]. Prior research has shown substantial evidence of this pattern in the case of Black male/White female marriages in the US, but weaker evidence for other types of unions [@gullickson_education_2006; @fu_racial_2001; @kalmijn_educational_2010; @hou_interracial_2011].

Regardless of its overall prevalence, the net effect of status exchange on resources is complex because status exchange positively selects one partner and negatively selects the other partner. Therefore the two effects may somewhat cancel each other out in terms of the overall distribution of resources across parents. Complicating this issue further, research has shown that the parents of biracial children may differ from parents of monoracial children in the magnitude and types of resources that they allocate to children [@cheng_constraints_2007], even among parents of similar resources.

Regardless of the underlying complexity of this selection process, delineating its role as opposed to other mechanisms is key to understanding the overall process that sorts biracial individuals within the racial hierarchy. Although this approach is sometimes characterized as pitting race versus class and minimizing the importance of race, that is not the case here. Instead, this delineation allows us to understand how race affects the outcomes of biracial individuals in a multigenerational process.

@@

The experience of hybridity itself may also positively or negatively affect outcomes for biracial individuals. The "marginal man" theory argues that biracial individuals will find it difficult to fit in due to the dissonance of an identity that crosses strong and salient racial boundaries [@park_human_1928; @stonequist_problem_1935]. This liminal identity conflict will then result in negative mental health experiences that could feed into other negative outcomes as well. Evidence for such negative mental health experiences, however, has been mixed [@udry_health_2003; @campbell_what_2006; @cheng_multiracial_2009; @bratter_does_2011].

Marginal man theorists viewed culture in monolithic and static terms and thus to be trapped in between two cultures was seen as a difficult experience. Contemporary views of culture instead emphasize the often flexible, fragmented, and fluid use of cultural bits that comprise a "toolkit" [@swidler_culture_1986; @dimaggio_culture_1997]. From this perspective, biracial individuals may benefit from an enlarged cultural toolkit that allows them to navigate a variety of diverse social contexts [@shih_multiple_2019].

The complexity of the underlying mechanisms described above prevents simple and straightforward hypotheses regarding the position of biracial individuals. Different mechanisms may produce the same observed outcome and the weight of different mechanisms can also vary by groups as these groups may be governed by different regimes of classification. However, comparing outcomes across biracial groups may help us better understand the likely mechanisms that drive biracial classification and to disregard others as implausible. Accounting for the effect of parental resources on outcomes to the best of our ability is also key to reaching any conclusions about the placement of biracial individuals within the racial hierarchy.

@@

Prior work on the outcomes for biracial individuals has been hampered by several methodological issues. With good reason, researchers have been reluctant to use data in which multiracial or biracial status is determined solely by the self-identification of respondents. We know that far fewer people identify as biracial than could feasibly do so based on their parents' races <!-- TODO cite morning? morning and Goldstein? who? --> and that various selection effects operate on the decision to identify as biracial among those who could [@roth_end_2005a; @davenport_role_2016]. Furthermore, research on changes in racial reporting over time suggests that other status markers like education and income may influence how people identify themselves [@saperstein_racial_2012; @saperstein_mulatto_2013]. For example, a person with one Black and one White parent who is routinely treated by others as Black and discriminated against as Black may be more likely to identify themselves as exclusively Black than a similar person of Black and White ancestry who experiences less consistent discrimination. Thus, any analysis of outcomes for biracial individuals based on self-identification must deal with a serious potential for reverse causation in the observed differences between biracial individuals and their constituent groups [@bratter_multiracial_2018].

Instead of self-identification, researchers have generally relied upon survey data in which the racial identification of a respondent's biological parents are provided in some form or another. @campbell_multiracial_2009 used data from the National Longitudinal Survey of Adolescent to Adult Health (Add Health) to examine high school GPA, advanced placement in math, and four-year college enrollment among biracial and monoracial individuals identified by co-resident biological parents' races. @herman_blackwhiteother_2009 examines differences in GPA using data on high school students in California and Wisconsin in which the students reported their parents' races. Both of these studies find some support for an "in-between" status for many biracial groups, but are both hampered by a lack of statistical power due to small sample sizes. @kao_racial_1999 used data from the National Educational Longitudinal Study of 1988 to examine mathematics achievement scores and GPA for biracial students with Black and Asian parentage and finds that biracial Black students have outcomes similar to monoracial Black students, while biracial Asian students have outcomes more similar to monoracial White students. This study, however, is hampered by the fact that biracial students are identified by a discrepancy between their reported race and one parent's reported race, rather than by both biological parents' races.

Results in prior work tend to be imprecise and inconclusive, due to the sample size for most biracial groups. An alternative approach is to use large scale Census or American Community Survey (ACS) data to identify biracial children by the race of the biological parents in their household. This approach can generate much larger samples of biracial individuals, although it does exclude examination of children living in non-nuclear households. However, researchers then have a limited range of outcomes to explore because biracial individuals identified by this method are by definition children. This approach has been used previously to examine the likelihood of living in poverty [@bratter_multiracial_2013; @bratter_poverty_2013] and residential segregation [@ellis_agents_2012]. These studies provide valuable information on biracial childrens' lived experience, but do not directly measure their own outcomes. However, one important educational outcome is partially identifiable for children in Census and ACS data - grade retention. I now turn to a closer examination of this particular outcome.

# Grade Retention as an Outcome

Grade retention is the practice of having students repeat a grade due to poor academic performance. Rates of grade retention likely increased throughout the late Twentieth century [@frederick_have_2008] but have declined substantially since 2004 [@warren_patterns_2014].

Prior work on grade retention has failed to show a positive effect of grade retention on later educational outcomes, and has instead found evidence of "scarring" effects that lead to later negative educational outcomes, such as a higher dropout risk [@jimerson_metaanalysis_2001; @stearns_staying_2007; @andrew_scarring_2014; @hughes_effect_2017]. The risk of grade retention also varies substantially by race and class background, with Black and Latino students at particularly high risk of grade retention relative to other groups [@warren_patterns_2014].

@@

Ideally, I would restrict the sample only to children living in a household with two biological parents. However, the ACS only records the direct family relationship between the head of household and other members of the family. Since 2008, the ACS also distinguishes between biological children, adopted children, and stepchildren. Respondents can therefore be identified as biological children of another member of the household only if that member is the head of household or the biological child of the head of household. Determining whether the partner/spouse of that biological parent is the other biological parent is more difficult.

Prior work has used a variety of additional restrictions to limit the analysis to children who are more likely to be the biological children of both parents [@saenz_persistence_1995; @xie_racial_1997]. Consistent with this prior work, I restrict cases in two ways. First, I restrict the sample to cases where both parents were of a reasonable age at the birth of the child (aged 15-44 for mothers and aged 15-60 for fathers). Second, I restrict the sample to only those cases where the reported race and Hispanicity of the child is inclusive of at least one of the parents' races.

The final analytical sample consists of children aged 5-20 who are currently enrolled in the K-12 school system and live in a household with two parents who are both likely to be biological parents. This necessary sample restriction will bias the family structure of the analytical sample relative to the total population of K-12 schoolchildren. The result of this bias is well-known. Children living in two-parent nuclear households have greater educational success than other children as a result of material and social resources [@biblarz_family_1999]. Therefore, this sample restriction will bias overall estimates of grade retention downward. However, this bias is less problematic for the goal of this study, which is to estimate differences between groups. The sample restriction effectively eliminates differences in family structure across racial groups that may account for some observed differences in grade retention. However, the remaining differences observed here are likely to hold for the full population in the absence of strong racial differences in the effect of family structure on grade retention.

The race of children is calculated by a cross-tabulation of parents' races. A full cross-tabulation of the race/Hispanicity of each parent would produce far too many combined categories to be analytically useful. Therefore, I collapse each parent's race into the categories of White, Black, Asian, Latino, and Indigenous. These categories form the "ethnoracial pentagon" that is commonly used in popular practice and government tabulation to identify race in the US [@hollinger_postethnic_1995]. Parents are identified as Latino based on their response to the Hispanicity question, regardless of their response to the race question. Parents are identified as Indigenous if they responded as either American Indian/Alaska Native or as Pacific Islander. Because the goal of my analysis is to examine specifically the outcomes of "first-generation" biracial children, I exclude cases where at least one parent identified with multiple races.

The cross-tabulation of parent's races leads to ten distinct biracial categories. The total sample size for each of these biracial categories is shown in @tbl-descriptive, along with the sample size of the monoracial comparison groups. The sample sizes are quite large for most groups and far in excess of comparable categories in other available survey data.

@@

I calculate two confidence intervals for each biracial group. Each confidence interval is in comparison to one of the constituent monoracial groups. These confidence intervals are color-coded in @fig-example to indicate the reference group. I also only draw half-intervals in the direction of the monoracial group's point estimate. Overlap in these color corresponding bars and bands indicates that the difference between the biracial group and the monoracial group is not statistically significant at $p<.05$. For example, the yellow bar shown in @fig-example for biracial Black/White students does not quite overlap with the yellow band for White students, indicating that the point estimates for these two groups are statistically distinguishable at $p<.05$.

@@

# Acknowledgments {.appendix}

I am grateful to...

>>>>

---
title: Differences in the Risk of Grade Retention for Biracial and Monoracial Students in the US, 2010-2019
author: 
  - Aaron Gullickson, *University of Oregon*
abstract: test abstract
thanks: The corresponding author can be reached at aarong@uoregon.edu. Supplementary materials provided with this article include full model results upon which figures are based as well as any senstivity analysis described herein. All code and data for this project are available at ...
format:
  pdf:
    documentclass: scrartcl
    papersize: letter
    fig-height: 4
    fig-pos: t
    keep-tex: false
  docx:
    toc: false
bibliography: ../project.bib
execute: 
  eval: true
  echo: false
  warning: false
  error: false
prefer-html: true
editor: source
csl: american-sociological-association.csl
---

@@

The experiences of mixed race individuals will play an important role in the future of race in the US. A rise in interracial unions since the 1960s has resulted in a "biracial baby boom" of individuals whose parents cross racial lines [@root_racially_1992] and this demographic surge may have transformative effects on racial boundaries [@hochschild_creating_2012; @alba_great_2020]. While racial mixing is nothing new to US history [@morning_who_2000; @gullickson_choosing_2011], this biracial boom is notable for its size and has occurred in an environment in which historical norms of racial classification may be eroding.

The transformative potential of this biracial baby boom is contingent on both how these biracial individuals identify their own race and how they are sorted into a prevailing system of racial inequality. While extensive scholarship documents the identification and classification decisions made by or on behalf of multiracial individuals [@xie_racial_1997; @qian_options_2004; @roth_end_2005a; @brunsma_interracial_2005; @herman_forced_2004; @holloway_place_2009; @bratter_mother_2009; @davenport_role_2016; @liebler_boundaries_2016], only a smattering of research has documented how the life chances of biracial individuals compare to their monoracial peers [@kao_racial_1999; @campbell_multiracial_2009; @herman_blackwhiteother_2009]. In this article, I expand on this work by examining the risk of grade retention for K-12 students whose parents identify with different races.

By examining this outcome, I am able to overcome two important limitations that have hampered prior work on the life chances of biracial individuals. First, using self identification of biracial respondents to analyze outcomes is problematic. Prior research indicates that individuals whose parents belong to different races do not consistently identify with both races [@doyle_are_2007; @liebler_america_2017; @gullickson_racial_2019]. Furthermore, social status and social mobility can predict racial identification generally [@saperstein_racial_2012; @saperstein_mulatto_2013] and social status is related to the identification choices of biracial individuals specifically [@roth_end_2005a; @davenport_role_2016]. Thus, differences in outcomes among biracial individuals may in part produce differences in identification, rather than the other way around.

This problem of reverse causation can be eliminated when researchers have access to the racial identification of a respondent's biological parents. Even when such information is available, however, researchers run into a second limitation of sample size. The samples of biracial respondents in many data sources are too small to make reliable statistical inferences. This limitation is particularly severe for biracial individuals with two non-white parents.

Biracial children who reside with both biological parents can be identified in large scale data sources such as the Census. Researchers have previously utilized this feature to examine the classification choices made on behalf of these children [@xie_racial_1997; @roth_end_2005a]. Analyzing outcomes for such children is more difficult, for the simple reason that they do not have many measurable outcomes yet. However, grade retention is measurable and important even for young children. Being held back in school is a relatively common occurrence that has consequences for later educational transitions [@jimerson_metaanalysis_2001] and is disproportionately applied by race [@warren_patterns_2014]. Grade retention is one of the earliest stratifiers of life chances that an individual will experience.

In this article, I use data from the American Community Survey to examine the likelihood of being held back in school for biracial children in comparison to their monoracial peers. Pooling data from 2010 to 2019, I am able to draw a large enough sample to make reliable statistical inferences for most biracial groups, including those with two non-white parents. The results provide a window into the placement of biracial children within an existing system of racial inequality. In contrast to prevailing hypotheses that predict either a pattern of hypodescent or an in-between position for biracial individuals, I find a consistent pattern in which the risk of grade retention is similar to that of the lower risk constituent monoracial group. This result holds even after adjusting for differences in resources across respondents, despite the fact that most biracial groups are positively selected by parental resources relative to their monoracial group.

@@

I distinguish three mechanisms that help us understand where biracial individuals might be positioned among these possibilities. First, I consider arguments regarding racial discrimination at both an individual and systemic level. Second, I consider arguments regarding different distributions of resources across racial groups. Third, I consider arguments regarding how the nature of hybridity itself may affect outcomes.

@@

Discrimination and bias by race play an important role in overall levels of racial inequality [@quillian_new_2006; @pager_sociology_2008]. Discrimination need not be consciously recognized by actors, but may instead reflect cognitive biases and schema that individuals unknowingly rely upon and reinforce in social interactions [@quillian_new_2006]. Discrimination can also be systemic in nature when biases are built into and reinforced by organizations and institutional systems [@reskin_race_2012]. In our current "colorblind" regime [@bonilla-silva_racism_2006], such systems may be formally race neutral but produce disparate impacts by race [@pager_sociology_2008].

@@

The strongest historical precedent for understanding such classification in the US is the "one-drop rule" norm governing classification as Black. According to the one-drop rule, individuals of known Black ancestry are considered exclusively Black, regardless of other ancestries [@davis_who_1991]. @davis_who_1991 generalizes this practice into the concept of *hypodescent* in which individuals with mixed racial ancestry will be classified as members of the lower status group. A parallel pattern of *hyperdescent* implies that individuals of mixed racial ancestry will be classified as members of the higher status group. This pattern of hyperdescent has historically governed the experiences of individuals with American Indian ancestry [@snipp_american_1989; @wolfe_land_2001].

Although the hypodescent/hyperdescent paradigm is often used to frame the experiences of multiracial individuals, its generalizability and contemporary relevance is questionable. Hypodescent and hyperdescent are observed in two specific historical cases involving Black ancestry and American Indian ancestry, respectively. As @iverson_regimes_2022 note, these two cases could just as parsimoniously be explained by a "dominance" model which ranks ancestries by their tendencies to be supercessive or recessive in determining identification and classification.

The development of particular norms may also reflect the historical regime in which they were developed [@gullickson_choosing_2011; @iverson_regimes_2022]. In this case, neither the hypodescent/hyperdescent or dominance paradigm may help us to fully understand the experiences of those of Latino and Asian ancestry due to more recent processes of racialization. Furthermore, the historical norms for all groups may be transforming in the post-Civil Rights era.

If outcomes for biracial individuals all closer to one of their constituent groups, then we may have stronger evidence about such norms of classification. However, what if those outcomes fall in-between the two groups? Processes of classification and discrimination could still produce an in-between status for several reasons.

First, racial ambiguity in appearance may lead to more inconsistent classification into single race categories across observers. Even if norms consistently indicate a certain type of classification, correct identification by observers is a probabilistic phenomenon and likely to be more heterogeneous for biracial individuals. As a result, biracial individuals will experience a lower sum total of discrimination and bias affecting the lower status group in the case of hypodescent and will not reap the same rewards as a member of the higher status group in the case of hyperdescent.

Second, biracial people's experience of discrimination may be "softened" because a more ambiguous physical appearance provokes less racial antipathy. The US already has a long history of this kind of softening. Because of the one-drop rule, Black individuals vary substantially in skin tone and those individuals with a lighter skin tone have better outcomes [@keith_skin_1991; @hughes_significance_1990; @monk_skin_2014] and are viewed more favorably by both Blacks and Whites [@maddox_cognitive_2002a]. Thus, light skin moderates the stigma of Blackness even among monoracial individuals.

Third, biracial individuals may be classified into an ambiguous "middleman" minority position. This kind of middle tier status for individuals of mixed race is common in Latin America [@telles_race_2009] and @bonilla-silva_biracial_2004 has argued that such a system may be emerging in the contemporary US as well. In this case, many biracial groups may occupy a nebulous and ill-defined middleman or buffer class position in-between whites and darker-skinned non-whites, helping to cement a new kind of racial hierarchy.

@@

Racial inequality is at least partly explained by different distributions of material, cultural, and social resources across racial groups [@conley_being_1999]. On average, socioeconomic background (e.g. parental income, wealth, and education) differs by race and such socioeconomic background is consequential for outcomes. Other less materially defined resources also vary across groups such as cultural styles, social networks, and neighborhood quality. These differences in background resources reflect the accumulation of historical processes of discrimination and, when combined with contemporary processes of discrimination, magnify the overall level of racial inequality we observe.

How do differences in the distribution of resources affect biracial individuals? Naively, we might expect biracial individuals to inherit parental resources roughly halfway between their two constituent groups, on average [@chew_american_1989]. Prior research looking at biracial children in the Census data has found some evidence to support this argument [@chew_american_1989]. In such a case, the resources of biracial individuals would predict outcomes in-between the two constituent groups, barring other mechanisms.

However, the actual resource position of biracial individuals is likely far more complicated because people are not randomly selected into interracial unions. The relationship between this selection process and a biracial child's outcomes depends on the strength and direction of that selection on each partner, its consistency across different types of unions, and how strongly the selected characteristic predicts outcomes in the next generation.

Status exchange theory predicts that given an unequal racial hierarchy, individuals from a lower status racial group will be positively selected on socioeconomic characteristics into intermarriage with a higher status group, while individuals from a higher status racial group will be similarly negatively selected into the same marriage [@davis_intermarriage_1941; @merton_intermarriage_1941; @fu_racial_2001]. Prior research has shown substantial evidence of this pattern in the case of Black male/White female marriages in the US, but weaker evidence for other types of unions [@gullickson_education_2006; @fu_racial_2001; @kalmijn_educational_2010; @hou_interracial_2011]. Even in cases where status exchange is present, its net effect on resources is complex because status exchange positively selects one partner and negatively selects the other partner. Therefore the two effects may somewhat cancel each other out in terms of the overall distribution of resources across parents.

Regardless of the underlying complexity of this selection process, delineating its role as opposed to other mechanisms is key to understanding the overall process that sorts biracial individuals within the racial hierarchy. Although this approach is sometimes characterized as pitting race versus class, that is not the case here. Instead, this delineation allows us to understand how race affects the outcomes of biracial individuals in a multigenerational process.

@@

The experience of hybridity itself may also positively or negatively affect outcomes for biracial individuals. According to "marginal man" theory, biracial individuals will find it difficult to fit in due to the dissonance of an identity that crosses strong and salient racial boundaries [@park_human_1928; @stonequist_problem_1935]. This liminal identity conflict will result in negative mental health experiences that could feed into other negative outcomes as well. Evidence for such negative mental health experiences, however, has been mixed [@udry_health_2003; @campbell_what_2006; @cheng_multiracial_2009; @bratter_does_2011].

Marginal man theorists viewed culture in monolithic and static terms and thus to be trapped in between two cultures was seen as a difficult experience. Contemporary views of culture instead emphasize the flexible, fragmented, and fluid use of cultural bits that comprise a "toolkit" [@swidler_culture_1986; @dimaggio_culture_1997]. From this perspective, biracial individuals may actually benefit from an enlarged cultural toolkit that allows them to navigate a variety of diverse social contexts [@shih_multiple_2019].

The complexity of the underlying mechanisms described above prevents simple and straightforward hypotheses regarding the position of biracial individuals. Different mechanisms may produce the same observed outcome and the weight of different mechanisms can also vary by groups as these groups may be governed by different regimes of classification. However, comparing outcomes across biracial groups may help us better understand the likely mechanisms involved and to disregard others as implausible. Accounting for the effect of parental resources on outcomes to the best of our ability is also key to reaching any conclusions about the placement of biracial individuals within the racial hierarchy.

@@

Prior work on the outcomes for biracial individuals has been hampered by several methodological issues. With good reason, researchers have been reluctant to use data in which multiracial or biracial status is determined solely by the self-identification of respondents. We know that far fewer people identify as biracial than could feasibly do so based on their parents' races [@morning_generational_2018] and that various selection effects operate on the decision to identify as biracial among those who could [@roth_end_2005a; @davenport_role_2016]. Furthermore, research on changes in racial reporting over time suggests that other status markers like education and income may influence how people identify themselves [@saperstein_racial_2012; @saperstein_mulatto_2013]. For example, a person with one Black and one White parent who is routinely treated by others as Black and discriminated against as Black may be more likely to identify themselves as exclusively Black than a similar person of Black and White ancestry who experiences less consistent discrimination. Thus, any analysis of outcomes for biracial individuals based on self-identification must deal with the potential for reverse causation in the observed differences between biracial individuals and their constituent groups [@bratter_multiracial_2018].

Instead of self-identification, researchers have generally relied upon survey data in which the racial identification of a respondent's biological parents are provided. @campbell_multiracial_2009 used data from the National Longitudinal Survey of Adolescent to Adult Health (Add Health) to examine high school GPA, advanced placement in math, and four-year college enrollment among biracial and monoracial individuals identified by co-resident biological parents' races. @herman_blackwhiteother_2009 examined differences in GPA using data on high school students in California and Wisconsin in which the students reported their parents' races. Both of these studies found some support for an "in-between" status for many biracial groups, but are hampered by a lack of statistical power due to small sample sizes. @kao_racial_1999 used data from the National Educational Longitudinal Study of 1988 to examine mathematics achievement scores and GPA for biracial students with Black and Asian parentage and finds that biracial Black students have outcomes similar to monoracial Black students, while biracial Asian students have outcomes more similar to monoracial White students. This study, however, is limited by the fact that biracial students are identified by a discrepancy between their reported race and one parent's reported race, rather than by both biological parents' races.

Results in prior work tend to be inconclusive, due to the sample size for most biracial groups. An alternative approach is to use large scale Census or American Community Survey (ACS) data to identify biracial children by the race of the biological parents in their household. This approach can generate much larger samples of biracial individuals. However, researchers then have a limited range of outcomes to explore because biracial individuals identified by this method are by definition children. This approach has been used previously to examine the likelihood of living in poverty [@bratter_multiracial_2013; @bratter_poverty_2013] and residential segregation [@ellis_agents_2012]. These studies provide valuable information on biracial childrens' lived experience, but do not directly measure their own outcomes. However, one important educational outcome is partially identifiable for children in Census and ACS data - grade retention.

Grade retention is the practice of having students repeat a grade due to poor academic performance. Prior work on grade retention has failed to show a positive effect of grade retention on later educational outcomes, and has instead found evidence of "scarring" effects that lead to later negative educational outcomes, such as a higher dropout risk [@jimerson_metaanalysis_2001; @stearns_staying_2007; @andrew_scarring_2014; @hughes_effect_2017]. The risk of grade retention also varies substantially by race and class background, with Black and Latino students at particularly high risk of grade retention relative to other groups [@warren_patterns_2014].

@@

Ideally, I would restrict the sample to children living in a household with two biological parents. However, while the ACS distinguishes between biological, adopted, and stepchildren, it only records the direct family relationship between the head of household and other members of the household. Respondents can therefore be identified as biological children of another member of the household only if that member is the head of household or the biological child of the head of household. Determining whether the partner/spouse of that biological parent is the other biological parent is more difficult.

Prior work has used a variety of additional restrictions to limit the analysis to children who are more likely to be the biological children of both parents [@saenz_persistence_1995; @xie_racial_1997]. Consistent with this prior work, I restrict cases in two ways. First, I restrict the sample to cases where both parents were of a reasonable age at the birth of the child (aged 15-44 for mothers and aged 15-60 for fathers). Second, I restrict the sample to those cases where the reported race and Hispanicity of the child is inclusive of at least one of the parents' races.

The final analytical sample consists of children aged 5-20 who are currently enrolled in the K-12 school system and live in a household with two parents who are both likely to be biological parents. This necessary sample restriction will bias the family structure of the analytical sample relative to the total population of K-12 schoolchildren. Children living in two-parent nuclear households have greater educational success than other children as a result of material and social resources [@biblarz_family_1999]. Therefore, this sample restriction will bias overall estimates of grade retention downward. However, this bias is less problematic for the goal of this study, which is to estimate differences between groups. The sample restriction effectively eliminates differences in family structure across racial groups that may account for some observed differences in grade retention. However, the remaining differences observed here are likely to hold for the full population in the absence of strong racial differences in the effect of family structure on grade retention.

The race of children is calculated by a cross-tabulation of parents' races. To produce reasonably parsimonious categories, I collapse each parent's race into the categories of White, Black, Asian, Latino, and Indigenous. These categories form the "ethnoracial pentagon" that is commonly used in popular practice and government tabulation to identify race in the US [@hollinger_postethnic_1995]. Parents are identified as Latino based on their response to the Hispanicity question, regardless of their response to the race question. Parents are identified as Indigenous if they responded as either American Indian/Alaska Native or as Pacific Islander. Because the goal of my analysis is to examine specifically the outcomes of "first-generation" biracial children, I exclude cases where at least one parent identified with multiple races. The cross-tabulation of parents' races leads to ten distinct biracial categories. The total sample size for each of these biracial categories is shown in @tbl-descriptive, along with the sample size of the monoracial comparison groups.

@@

I calculate two confidence intervals for each biracial group. Each confidence interval is in comparison to one of the constituent monoracial groups. These confidence intervals are color-coded in @fig-example to indicate the reference group. I only draw half-intervals in the direction of the monoracial group's point estimate. Overlap in these color corresponding bars and bands indicates that the difference between the biracial group and the monoracial group is not statistically significant at $p<.05$. For example, the yellow bar shown in @fig-example for biracial Black/White students does not quite overlap with the yellow band for White students, indicating that the point estimates for these two groups are statistically distinguishable at $p<.05$.

@@
